A multi-scale story of the diffusion of a new technology: the web

### 2024-09-12 paper/submission1/tranos2023.qmd
postpresentation2

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol and The Alan Turing Institute
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  This paper maps the participation in the digital economy and its evolution in the UK over space and time. Most of the existing economic geography literature which dealt with the spatiality of the internet employed supply-side measures, such as infrastructural capacity, in order to understand the geography of the digital economy and its potential spatial economic effects. Useful as these approaches might have been, they cannot capture the micro-processes and the characteristics of the individual online behaviour. Using large volumes of archived and geolocated web content, this paper models the diffusion of web technologies over space and time in the UK. The data and geolocation strategy allow to capture these processes at a very granular spatial scale. The modelling approach, which is based on simple spatial analytical methods and on the estimation of diffusion curves at various scales, enables to depict the role of geography and other cognitive factors which drove the diffusion of web technologies. Although the focus is on a recent historical period -- 1996-2012 -- the results of the analysis depict diffusion mechanisms which can be very useful in understanding the evolutionary patterns of the adoption of other newer technologies.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

#insert libraries here
library(rprojroot)
library(tidyverse)
library(knitr)
library(kableExtra)
library(sf)
library(cowplot)

@@

<!-- geographers lack of interest + motivation -->

Geographers were always interested in how new technologies and
innovations diffuse across space and time and, importantly, how such
spatio-temporal processes can be modeled. After all, diffusion together
with invention and innovation are considered the pillars of
technological change [@das2022diffusion]. The seminal contribution of
@hagerstrand1968innovation is illustrative of this early interest.
However, the torch of exploring and modelling such processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space
and time can be attributed to the scarcity of relevant and granular
enough data. As @zook2022mapping highlight, digital activities are
hardly ever captured in official data.

<!-- contribution -->

This paper offers such a contribution: a geographical study illustrating
how a new technology that is the web diffused over space and time in the
UK at a high level of spatial granularity during the $1996-2012$ period.
It does so by employing a novel source of big data which captures the
active engagement with web technologies during that period. By
addressing this empirical question this paper exemplifies how the
combination of data sources which escape the traditional social science
domain and adequate research methods can offer new lenses to
geographical research regarding the understanding of technological
diffusion.

**SAY SOMETHING ABOUT SPATIAL ANALYTICS, DIST PAPER BY ARP, NOT SPECIFIC EXPLANATORY VARIABLES, INTANGIBLE**

<!-- motivation -->

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related *future* technologies. As per
@leibowicz2016representing, historical studies agree that technologies diffuse 
differently in terms of times, rates, and geographies and can be driven
by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers; network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there are very
limited attempts in the literature to analyse the diffusion of new
digital technologies at a detailed geographical level.

<!-- diffusion and adoption -->

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, @rogers2010diffusion identifies early adopter of new
technologies as 'knowledgeable risk takers' and @griliches1957 as
'profit maximisers' [@ding2010modeling]. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, @perkins2011internet attribute diffusion
to two processes: (i) epidemic-like mechanisms, which are governed by
distance, proximity and social interactions, and (ii) economic
mechanisms as new innovations are adopted by users as they become more
profitable, valueable and useful.
<!-- The criticisms are mainly based on the fact that although the approach -->
<!-- gives an idea of aggregate (industry or household) behaviour, it does -->
<!-- not focus on the individual's (firm or household) adoption process -->

<!-- data and early engagement with the web -->

This paper focuses on the diffusion of the web as new technology during
the $1996-2012$ period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
$9$ per cent of UK's households had access to the internet in $1998$
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing was happening exclusively from desktop PCs
as there were no smartphones [@tranosuk]. Hence, it is fair to say that
the study period captures the very early stages of the diffusion on a
new technology that is the web as well as its maturity. 
The former is a key point in the lifecycle of a
new technology. Firstly, during this period new technologies are
expensive, crude and imperfect [@rosenberg1994exploring; @wilson201281].
A simple comparison between Web 1.0 and Web 3.0 applications clearly
illustrates this [@tranos2020social]; for instance a static website compared 
with a platform like `github`, which enables cooperation between users and 
the creation of new information, meaning, and knowledge 
[@faraj2016special; @barassi2012does]. 
During this period the performance
of a new technology is the main attraction and not the cost to access
and use it [@wilson2011lessons]. There is a broader theoretical
discussion in the literature regarding the motivation behind early
adoption. As summarised by @perkins2005international, on the one hand,
epidemic models highlight the role of interpersonal contacts as a way
for new technologies to diffuse. On the other hand, economic models
underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies because 
of the commonly found lag between investment and economic returns as reflected in 
the Solow paradox [@acemoglu2014return; @brynjolfsson2018artificial].

@@

Importantly, the data used here depicts the *active* engagement with the
web in the UK as it contains all commercial websites that 
(i) are part of the UK's relevant second level domain (SLD, .co.uk), 
(ii) have been archived by the Internet Archive[^2], and 
(iii) include a mention to at least one valid UK
postcode in the web text. 
The act of creating a website is understood here as active engagement with the
web vis-à-vis the more passive act of browsing the web or having an
internet connection [@tranosuk]. Previous studies have focused mostly on
more passive notions of engaging with digital technologies such as
internet adoption and internet speeds [e.g. @blank2018local;
@riddlesden2014broadband]. More details about the data and the data
generation process can be found in @sec-datamethods.

**S curve??**

<!-- description of the methods and results -->

<!-- contents -->

@grubler1990rise Later Hagerstrand conceptualized physical "barrier"
effects like lakes or uninhabited areas, which, in addition to distance,
act as further retarding effects on diffusion. These are formalized in
the form of "zero" or "half" contact multiplicators on the (distance
decaying) message flows.

@grubler1990rise With respect to the formalization of the communication
flows Hagerstrand defines a "mean information field" (MIF), in which the
probability of communication is a negative function of distance between
individuals

@wilson201281 Logistic growth describes an initial period of gradual
diffusion as a technology is introduced as a new commercial application,
moving then through a rapid, exponential growth phase, before slowing
and eventually saturating [@grubler1999dynamics]. The substitution of
incumbent technologies by new competitors leads to subsequent decline
and eventual obsolescence.

@@

<!-- key concepts -- copied from help.Rmd -->

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be identified. 
Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, there is a hierarchical processes. Instead of
horizontal distance-based diffusion mechanisms, the top-down hierarchy
of urban systems shapes technological diffusion. In reality, the synthesis of these
two processes represents how new technologies diffuse over space and
time [@morrill2020spatial].

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled 'Innovation Diffusion as a Spatial Process'
[@hagerstrand1968innovation]. Hägerstrand was the first one to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified
the 'neighbourhood' effect of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the 'lead-lag'
spatial acceleration effect in market research [@bento2018time;
@PERES201091]. Hägerstrand is more widely known though for highlighting
the role time plays in the diffusion of innovations: an early-pioneering
period, a middle fast accelerating period and a final saturation one
[@morrill2020spatial].

The temporal dimension was further explored by Everett Rogers and his
seminal work on 'Diffusion of Innovations' [@rogers2010diffusion].
Rogers being a sociologist focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism of
diffusion and adoption is communication and how knowledge is
transferred within a social system. Therefore, all approaches agree on
the S-shaped diffusion and cumulative adoption pattern
[@grubler1990rise].

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing]. @grubler1990rise effectively
summarises the three key characteristics of the spatial diffusion
process: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy effect in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the three
mechanisms that the empirical analysis in @sec-results investigates.

<!-- previous studies -->

The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted
here. @beardsell1999spatial studied the evolution of the computer
industry in $317$ US metro areas during the $1977-1992$ period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across $402$ German regions during $1970-2015$. Their
key finding is that local demand than local supply was the main driving
factor.

At a global scale @perkins2005international explored whther the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
@perkins2011internet explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones was
shaped by similar socioeconomic factors as the internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, @ding2010modeling modelled the spatial diffusion of
mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network, @lengyel2020role
analysed its adoption and the churn at a very granular spatial level.
Their results are in agreement with early theoretical and empirical
contributions reviewed here: assortativity, urban scaling and distance
are the key drivers of spatial diffusion. At a global scale
@PAPAGIANNIDIS2015308 modelled the diffusion of different web
technologies technologies and practices. Interestingly, they did so by
using similar, but less extensive data as the one used here. Their
analysis illustrated how the diffusion of different web technologies and
practices follow an S-shaped pattern as well as the different diffusion
rates of the different technologies and practices.

All in all, ...

@@

<!-- @bento2018time explore What determines the duration of formative phases -->

<!-- for energy innovations in different markets? We are interested both in -->

<!-- initial markets (also: core, lead, first mover, early adopter) where -->

<!-- formative phases prepare technologies for mass commercialization, and in -->

<!-- follower markets (also: periphery, lag, late adopter) where accelerated -->

<!-- formative phases may benefit from diffusion and spillovers. -->

@@

**ADD FAMILY OF WEB TECHNOLOGIES**

<!-- main metric -->

To capture the diffusion of web technologies, a website density metric is developed 
for two different geographical scales: the Local Authority Districts (LAD) and the 
Output Areas (OA). The former is an administrative unit and there are 
c. 374 such units in the UK. The latter is a census-based geographical unit, which 
is very small as there are c. 230,000 of them in the UK. This methodological choice 
will allow the mapping of the diffusion of web technologies and the assessment of 
the diffusion mechanisms at these two very different spatial scales.

<!-- web data -->

The counts of websites at these scales are calculated using data from the 
Internet Archive [^2] and, specifically, the JISC UK Web Domain Dataset [@ukwebarchive]. 
The Internet Archive is one of the most complete and oldest archive of webpages 
in the world operating since 1996 [@ainsworth2011much; @holzmann2016dawn].
It is a web crawler, which discovers webpages by following the hyperlinks
of every webpage its archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK ccTLD (.uk) from the 1996–2012 period. 
In essence, this is a long list of 2.5 billion URLs of archived webpages including 
also the archival timestamp.

@@

Instead of using the whole .uk country code top-level domain (ccTLD), this 
paper focuses on its commercial subset, the .co.uk SLD. 
This choice decrease the heterogeneity of the web data as 
such commercial websites have rather specific aims: they are used to diffuse information,
support online transactions and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as .com and these cases escape 
the data used here, such omissions should not affect our results given the
popularity of the .uk ccTLD [@tranosuk]: UK consumers prefer to visit a .uk website
when they are searching for products or services [@hope]; and
anecdotal evidence indicates that during the first half of 2000, three
.co.uk domains were registered every minute [@oecd_coms]. 
Importantly, previous studies illustrated that .co.uk is the most popular UK SLD
[@tranosuk].

@@

The data cleaning process included an aggregation step, through which
the archived webpages were aggregated to the parent websites. This website 
reconstruction allows the creation of a *website* instead of a *webpage* density 
metric. Websites, which are hierarchical structures, tend to represent specific 
orgnanisations or entities, and, arguably, are more meaningful observational units 
than webpages, which ignore the upstream dependencies. 
Based on the following example, all three webpages are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)) and at the end 
only websites and not the nested webpages were considered as otherwise the metric 
would have been biased towards large, place specific, websites.

@@

1.  S-shaped diffusion curves: S for LAD per firm. UK,
    fast/slow/examples

2.  ranks: there is stability and movement

>>>>

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  This paper maps the participation in the digital economy and its evolution in the UK over space and time. Most of the existing economic geography literature which dealt with the spatiality of the internet employed supply-side measures, such as infrastructural capacity, in order to understand the geography of the digital economy and its potential spatial economic effects. Useful as these approaches might have been, they cannot capture the micro-processes and the characteristics of the individual online behaviour. Using large volumes of archived and geolocated web content, this paper models the diffusion of web technologies over space and time in the UK. The data and geolocation strategy allow to capture these processes at a very granular spatial scale. The modelling approach, which is based on simple spatial analytical methods and on the estimation of diffusion curves at various scales, enables to depict the role of geography and other cognitive factors which drove the diffusion of web technologies. Although the focus is on a recent historical period -- 1996-2012 -- the results of the analysis depict diffusion mechanisms which can be very useful in understanding the evolutionary patterns of the adoption of other newer technologies.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

#insert libraries here
library(rprojroot)
library(tidyverse)
library(knitr)
library(kableExtra)
library(sf)
library(cowplot)
library(patchwork)

@@

<!-- geographers lack of interest for diffusion + motivation -->
Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of @hagerstrand1968innovation,
which demonstrated the importance of communications as a mechanism for the 
diffusion of innovations, being the most impactful. It was followed by various 
empirical studies -- see for instance @ormrod1990 and @iso2005. 
However, the torch of exploring and modelling such diffusion processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the recent lack of
geographical studies exploring the diffusion of new and, more specifically for 
this paper, digital technologies across both space and time can be attributed to 
the scarcity of relevant and granular enough data, a problem also highlighted 
by @iso2005 and @kemeny2011international. As @zook2022mapping pointed, digital 
activities are hardly ever captured in official data.

<!-- contribution -->
This paper offers such a contribution: a geographical study demonstrating
how a new technology that is the Web diffused over space and time in the
UK at a high level of spatial granularity during the 1996-2012 period.
It does so by employing a novel source of big data and machine 
learning algorithms to model the *active* engagement with Web during that 
period and tests how different spatial and temporal mechanisms shaped its 
diffusion. This paper also exemplifies how the combination of data sources 
which escape the traditional social science domain and adequate research 
methods can offer new lenses to geographical research regarding the 
understanding of technological diffusion.

<!-- motivation: stakeholders -->
The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
about the diffusion of related *future* technologies. As per
@leibowicz2016representing, historical studies agree that technologies diffuse 
differently in terms of times, rates, and geographies and can be driven
by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers, network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there is a scarcity 
of empirical studies analysing the diffusion of new digital technologies 
at a detailed geographical level.

<!-- motivation: regional development -->
This gap becomes even more prominent if we consider the importance of 
technology and its adoption in economic geography and macroeconomic growth 
theories. Simply put, differences in technological adoption and sophistication 
may lead to uneven development as new technologies can increase productivity 
and allow firms to introduce new products [@solow1957technical; 
@aghion1990model; @kemeny2011international]. Linked to the subject of this 
paper, the Web -- and the active engagement with this technology that is 
creating instead of just browsing a website -- can be considered as the 
underpinning technology for the transformative process known as the
*digital service economy* [@capello2024nexus].
After all, diffusion together with invention and innovation are considered 
the pillars of technological change [@das2022diffusion].
Findings from previous empirical work illustrated that early engagement with 
digital technologies such as the Web led to longstanding positive productivity 
gains for these regions that hosted early adopters [@tranosuk].

<!-- motivation: intangible/digital -->
Another line of motivation has to do with the intangible nature of the Web 
and the debate about the footloose nature of digital technologies, which
contrary to several earlier general purpose technologies are light in weight 
[@coyle1999weightless]. Following the seminal work of @gaspar1998information, 
which demonstrated that technological improvements in telecommunications lead 
to increased demand for face-to-face communications and, consequently, 
enhance the importance of cities, empirical studies from economic geography 
and urban economics explored whether digital technologies substitute or 
complement agglomeration externalities and spatial structure 
[@kolko_death_2000; @ioannides2008effect; @tranos2021ubiquitous].
With the exception of @sinai2004geography, who focused, among other things, 
on web content and demonstrated that more such content is available in larger
markets, this strand of literature has been mostly concerned with the
spatiality of the internet's hard and tangible infrastructure
and shied away from exploring the spatial dimensions of one of the 
most distinct software layers of the internet infrastructure, the Web.

<!-- diffusion and adoption -->
Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, @rogers2010diffusion identifies early adopter of new
technologies as 'knowledgeable risk takers' and @griliches1957 as
'profit maximisers' [@ding2010modeling]. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, @perkins2011internet attribute diffusion
to two processes: (i) epidemic-like mechanisms, which are governed by
distance, proximity and social interactions, and (ii) economic
mechanisms as new innovations are adopted by users as they become more
profitable, valueable and useful.
<!-- The criticisms are mainly based on the fact that although the approach -->
<!-- gives an idea of aggregate (industry or household) behaviour, it does -->
<!-- not focus on the individual's (firm or household) adoption process -->

<!-- early engagement with the web -->
This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
9 per cent of UK's households had access to the internet in 1998
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing involved exclusively a desktop PCs
as there were no smartphones [@tranosuk]. Hence, it is fair to say that
the study period captures the very early stages of the diffusion of a
new technology that is the Web as well until its maturity. 
The former is a key point in the lifecycle of a new technology. 
Firstly, during this period new technologies are expensive, crude and 
imperfect [@rosenberg1994exploring; @wilson201281].
A simple comparison between Web 1.0 and Web 3.0 applications clearly
illustrates this [@tranos2020social]: while a static website facilitates
one-way information dissemination, a platform like `github` enables 
cooperation between users and the creation of new information, meaning, 
and even knowledge [@faraj2016special; @barassi2012does]. 
During this period the performance of a new technology is the main 
attraction and not the cost to access and use it [@wilson2011lessons]. 
There is a broader theoretical discussion in the literature about 
the early adoption motives. As summarised by @perkins2005international, 
on one hand, epidemic models highlight the role of interpersonal 
contacts as a way for new technologies to diffuse. On the other hand, 
economic models underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies because 
of the commonly found lag between investment and economic returns as reflected in 
the Solow paradox [@acemoglu2014return; @brynjolfsson2018artificial].

@@

<!-- data -->
The data used here depict the active engagement with the Web as they contain 
geolocated and time-stamped counts of commercial websites from the UK. Instead 
of adopting metrics reflecting the passive engagement with the internet such as 
internet adoption and speeds [e.g. @blank2018local; @destefano2022fuzzy], this 
paper observes yearly counts of geolocated websites, which reflect the creation 
and maintenance of websites. It needs to be highlighted here that although 
there are various types of underpinning web technologies -- namely markup, 
stylesheet, client- and server-side coding languages as well as different
web browsers and other related programming language -- the focus here
is on the diffusion of the Web in its totality. For a more detailed and 
cross-country technological diffusion study, see @PAPAGIANNIDIS2015308.

<!-- description of the methods and results -->
To analyse these data, a diverse set of methods are employed. Firstly, 
cumulative adoption curves are estimated for the UK as a whole and for the Local 
Authority Districts (LAD), illustrating, for the first time, such S-shaped 
diffusion patterns for local areas. This system level analysis also reveals 
insights about the volatility of the adoption of the Web. Then, exploratory 
spatial data analysis offers some first indications of the importance of the 
different spatial diffusion mechanisms in shaping the diffusion of the Web in 
the UK. Importantly, the analysis adopts two distinct geographical scales: 
the LAD and the much more detailed scale of the smallest census areas in the UK,
the Output Areas. Lastly, a novel methodological framework based on 
well-established Machine Learning (ML) algorithms is developed to test the role 
of such mechanisms at these two scales. The results reveal for the first time 
the importance and relevance of different spatial diffusion mechanisms for the 
diffusion of a tangible, digital technology at very granular geographical scales.

<!-- contents -->
The structure of the paper goes as follows. @sec-litreview reviews the key
concepts in the diffusion literature as well as relevant empirical studies. 
Then, @sec-datamethods describes the data and methods used in this paper. 
The results of the different stages of the analysis are presented in 
@sec-results. The discussion of the results and the conlcusions of the paper
are offered in @sec-conclusions.

<!-- @grubler1990rise Later Hagerstrand conceptualized physical "barrier" -->
<!-- effects like lakes or uninhabited areas, which, in addition to distance, -->
<!-- act as further retarding effects on diffusion. These are formalized in -->
<!-- the form of "zero" or "half" contact multiplicators on the (distance -->
<!-- decaying) message flows. -->

<!-- @grubler1990rise With respect to the formalization of the communication -->
<!-- flows Hagerstrand defines a "mean information field" (MIF), in which the -->
<!-- probability of communication is a negative function of distance between -->
<!-- individuals -->

**SAY SOMETHING ABOUT DIST PAPER BY ARP, NOT SPECIFIC EXPLANATORY VARIABLES**

@@

<!-- key concepts -- copied from help.Rmd -->
Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be 
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, there is a hierarchical processes. Instead of a
horizontal distance-based diffusion mechanism, the top-down hierarchy
of urban systems shapes technological diffusion. In reality, the synthesis 
of these two processes better reflects how new technologies diffuse over 
space and time [@morrill2020spatial].

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled 'Innovation Diffusion as a Spatial Process'
[@hagerstrand1968innovation]. Hägerstrand was the first one to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified
a 'neighbourhood' mechanism of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the 'lead-lag'
spatial acceleration effect in market research [@bento2018time;
@PERES201091]. Hägerstrand is more widely known though for highlighting
the role time plays in the diffusion of innovations: an early-pioneering
period, a middle fast accelerating period and a final saturation one
[@morrill2020spatial].

The temporal dimension was further explored by Everett Rogers and his
seminal work on 'Diffusion of Innovations' [@rogers2010diffusion].
Rogers being a sociologist focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism of
diffusion and adoption is communication and how knowledge is transferred 
within a social system. All above approaches agree that diffusion and the
cumulative adoption of a new technology are represented by an S-shaped 
line plot [@grubler1990rise].

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing].

@grubler1990rise effectively summarises the three key spatial diffusion 
mechanisms: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy mechanisms in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and proximity. 
These are the three mechanisms that the empirical analysis in @sec-results 
investigates.

<!-- previous studies -->
The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metro areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modeled their
spatial diffusion across 402 German regions during 1970-2015. Their
key finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a
region or industry with high share of firms having a website, they are
more likely to have a website too.

At a global scale @perkins2005international explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
@perkins2011internet explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, @ding2010modeling modeled the spatial diffusion of
mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network, @lengyel2020role
analysed its adoption and the churn at a very granular spatial level.
Their results are in agreement with early theoretical and empirical
contributions reviewed here: assortativity, urban scaling and distance
are the key drivers of spatial diffusion. @bakher2013diffusion compared 
different the growth curves as a means to model diffusion of online 
shopping in Australia during the 1998-2009 period. At a global scale
@PAPAGIANNIDIS2015308 modeled the diffusion of different web
technologies and practices across countries. Interestingly, they used 
similar, but less extensive data as the ones used here. Their analysis 
illustrated how the diffusion of different web technologies and
practices follow an S-shaped pattern as well as the different diffusion
rates of the different technologies and practices.

All in all, this paper tests whether the three distinct spatial diffusion 
mechanisms shaped the diffusion of the Web in the UK at very granular
spatial scales from the Web's early stages onwards.

@@

<!-- @bento2018time explore What determines the duration of formative phases -->
<!-- for energy innovations in different markets? We are interested both in -->
<!-- initial markets (also: core, lead, first mover, early adopter) where -->
<!-- formative phases prepare technologies for mass commercialization, and in -->
<!-- follower markets (also: periphery, lag, late adopter) where accelerated -->
<!-- formative phases may benefit from diffusion and spillovers. -->

@@

<!-- main metric -->
To capture the diffusion of the Web, a website density metric is developed 
for two different geographical scales: the Local Authority Districts (LAD) and the 
Output Areas (OA). The former is an administrative unit and there are 
374 LAD in the UK. The latter is the smallest census-based geographical unit and 
there are c. 230,000 of them in the UK. This methodological choice 
will allow the mapping of the diffusion of web technologies and the assessment of 
the diffusion mechanisms at these two very different spatial scales.

<!-- web data -->
The counts of websites at these scales are calculated using data from the 
Internet Archive[^2] and, specifically, the JISC UK Web Domain Dataset 
[@ukwebarchive]. The Internet Archive is one of the most complete and oldest 
archive of webpages in the world operating since 1996 
[@ainsworth2011much; @holzmann2016dawn]. It is a web crawler, which discovers 
webpages by following the hyperlinks of every webpage it archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK ccTLD (.uk) from the 1996–2012 period. 
In essence, this is a long list of 2.5 billion URLs of archived webpages including 
also the archival timestamp.

@@

Instead of using the whole .uk ccTLD, this paper focuses on its commercial 
subset, the .co.uk second level domain (SLD). This choices decrease the 
heterogeneity of the web data as such commercial websites have specific 
aims: they are used to diffuse information, support online transactions 
and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as .com and these 
cases escape the data used here, such omissions should not affect the 
validity of the results given the popularity of the .uk ccTLD [@tranosuk]: 
UK consumers prefer to visit a .uk website when they are searching for 
products or services [@hope]; and canecdotal evidence indicates that during 
the first half of 2000, threec .co.uk domains were registered every minute 
[@oecd_coms]. Importantly, previous studies illustrated that .co.uk is the most 
popular UK SLD [@tranosuk].

@@

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This website 
reconstruction process lend itself to the consteruction of *website* instead 
of *webpage* density metrics. Websites tend to represent specific orgnanisations 
or entities, and, arguably, are more meaningful observational units than 
webpages, which can ignore the upstream dependency of the website they belong to. 
Based on the following example, all three webpages are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)). Only 
the overall website and not the nested webpages are considered for the counts 
as otherwise the density metrics would have been biased towards large, place 
specific, websites.

@@

## System-level analysis
